Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-16 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: cleaned dcs appendix

How can we determine the beyond-GR corrections? In GR, comparison between the numerically determined
Kerr QNMs against the fitting formula~\eqref{eq:qnm_nongr_iso} fixes the GR expansion coefficients
$\omega^{(n)}$, $\tau^{(n)}$.
%
We can proceed in a similar way with QNMs calculated in the context of a beyond-GR theory.
%
In particular, in the literature, we can already find fitting formulas relating
the QNMs to the BHs mass, spin and coupling constant $\ell$, the latter being
specific to each theory, up to $n=1$ in the spin expansion.
%
The idea is then to compare these formulas against Eq.~\eqref{eq:qnm_nongr_iso}
to fix $p$, $\delta\omega^{(n)}$, and $\delta\tau^{(n)}$.

@@

In Fig.~\ref{fig:example_waveform} we show an illustrative waveform for GR (solid line; using the \texttt{SEOBNR} model) and in the cubic EFT of gravity (dashed line; using the \pSEOB{} model),
including the leading-order $n = 0$ deformations to the fundamental QNM (see Table~\ref{tab:ref_theories_qnms}).
%
We chose binary parameters similar to GW150914, consisting of a quasicircular,
nonspinning binary with detector-frame masses
$m_1^{\rm dec} = 39\msun$ and $m_2^{\rm dec} = 31\msun$.
%
The top panel shows the $+$ polarization of the GW $h_{+}$ in both theories,
while the bottom panel shows the amplitude $|h| = \sqrt{h_{+}^2 + h_{\times}^2}$
and instantaneous frequency $f$, all as functions of time $t$.
%
The signal are identical up to the merger, after which they differ during the ringdown.
%
By construction, the ringdown is lasts longer for the cubic EFT of GR waveform and
with a smaller instantaneous frequency (see the bottom panel) due to the negative value of the
$\delta\omega^{(0)}$ coefficient in this theory.

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/example_waveform_cubicEFT.pdf}
\caption{(Color online).~Example of an waveform.}
\label{fig:example_waveform}
\end{figure}

@@

Assuming Gaussian and stationary noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln P(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}\,,
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk denotes
complex conjugation and $S_{n}(f)$ is the power spectrum density of the
detector.
%
Assuming a specific prior distribution for our parameters, we stochastically
sample over the parameter space using a Markov-Chain Monte Carlo algorithm as implemented in
\texttt{LALInferenceMCMC}~\cite{Rover:2006ni,vanderSluys:2008qx},
which is part of the \texttt{LALInference} software suite~\cite{Veitch:2014wba,lalsuite}.
%
We subsequently marginalize over the remaining parameters to obtain the
posterior probability distribution function~(PDF) on $\ell$.
%
An example of this is shown in Fig. XX.

@@

Let us show one concrete example of how this mapping is done.
%
In Ref.~\cite{Wagle:2021tam} the QNMs of slow-rotating BHs in dCS gravity were calculated
and it was found that the axial gravitational modes have their damping time increased as we
increase the Chern-Simons coupling $\ell_{\rm CS}$ at fixed spin $\chi$ of the BH.
%
Having selected this branch of QNMs we can proceed to determine $\delta\omega^{(i)}$ and $\delta\tau^{(i)}$ as follows.
%
Using the fitting formula Eq.~(54a) of~\cite{Wagle:2021tam}, namely,
%
\begin{equation}
    M \omega^{\rm CS}_{220} = c_{1} + c_{2} \kappa \zeta + (c_{3} + c_{4} \kappa \zeta) \, (1 - \chi)^{c_{5} + c_{6} \kappa \zeta},
    \label{eq:omega_dsc_eg}
\end{equation}
%
and similarly for the imaginary part ${\rm Im}(\sigma_{220}) =  - 1 / \tau^{\rm CS}_{220}$.
%
Here $\kappa = 1/(16 \pi)$, $\zeta = \ell_{\rm CS}^{4} / (M^4 \kappa)$ [thus $\kappa \zeta = (\ell_{\rm CS} / M)^4$]
and $c_{i}$ are fitting coefficients.
%
In the context of a binary BH merger, $M$ is interpreted as the remnant mass $\Mf$.

We can now expand Eq.~\eqref{eq:omega_dsc_eg} to leading-orders in $\chi$ and $\ell_{\rm CS}$, and
gather the terms proportional to $\ell_{\rm CS}$.
%
Doing so, we find,
%
\begin{align}
    \Mf \, \omega^{\rm CS}_{220} &=
    \left[ 0.3722 + 1.1945 (\ell_{\rm CS} / \Mf)^4 \right]
    \nonumber \\
    &\quad + \left[0.1861 + 5.1828 (\ell_{\rm CS} / \Mf)^4 \right] \, \chi\,,
\end{align}
%
where we made use of the numerical value of the coefficients $c_{i}$ and,
reassuringly, the nonrotating and linear-in-spin coefficients above agree with those
of~\cite{Maselli:2019mjd} to xx\%.
%
We now isolate the dCS corrections and compare against Eq.~\eqref{eq:qnm_nongr_iso},
which yields,
%
\begin{equation}
p_{\rm CS} = 4\,, \quad \delta \omega^{(0)}_{\rm CS} = 3.1964\,, \quad \delta \omega^{(1)}_{\rm CS} = 41.199\,.
\label{eq:cs_omega_coefs}
\end{equation}
%
We carry the same steps for $\tau^{(i)}$, which results in,
%
\begin{equation}
\delta \tau^{(0)}_{\rm CS} = 6.3619\,, \quad \delta \tau^{(1)}_{\rm CS} = 794.66\,.
\label{eq:cs_tau_coefs}
\end{equation}
%
which completes the set of fixed parameters beyond-GR parameters in the ringdown of the \pSEOB{} waveform model.
%
We remark that the alarmingly large values of $\delta \omega_{\rm CS}^{(1)}$ and $\delta \tau_{\rm CS}^{(1)}$
are compensated by the assumptions that $(\ell_{\rm CS} / M_f)^2$ and $\chi$ are much less than unity. These
are the assumptions use to calculate the QNM frequencies both in Ref.~\cite{Wagle:2021tam} and also in
other gravity theories considered here~\cite{Pierini:2021jxd,Cano:2021myl}.
%
As we said before, the only free beyond-GR parameter left is now $\ell_{\rm CS}$.

>>>>

How can we determine the beyond-GR corrections? In GR, comparison between the numerically determined
Kerr QNMs against the fitting formula~\eqref{eq:qnm_nongr_iso} fixes the GR expansion coefficients
$\omega^{(n)}$, $\tau^{(n)}$.
%
We can proceed in a similar way with QNMs calculated in the context of a beyond-GR theory.
%
In particular, in the literature, we can already find fitting formulas relating
the QNMs to the BHs mass, spin and coupling constant $\ell$, the latter being
specific to each theory, up to $n=1$ in the spin expansion (see Table~\ref{tab:ref_theories_qnms}).
%
The idea is then to compare these formulas against Eq.~\eqref{eq:qnm_nongr_iso}
to fix $p$, $\delta\omega^{(n)}$, and $\delta\tau^{(n)}$.

@@

In Fig.~\ref{fig:example_waveform} we show an illustrative waveform for GR (solid line; using the \texttt{SEOBNR} model) and in the cubic EFT of gravity (dashed line; using the \pSEOB{} model),
including the leading-order $n = 0$ deformations to the fundamental QNM (see Table~\ref{tab:ref_theories_qnms}).
%
We chose binary parameters similar to GW150914, consisting of a quasicircular,
nonspinning binary with detector-frame masses
$m_1^{\rm dec} = 39\msun$ and $m_2^{\rm dec} = 31\msun$.
%
The top panel shows the $+$ polarization of the GW $h_{+}$ in both theories,
while the bottom panel shows the amplitude $|h| = (h_{+}^2 + h_{\times}^2)^{1/2}$
and instantaneous frequency $f$, all as functions of time $t$.
%
The signal are identical up to the merger, after which they differ during the ringdown.
%
By construction, the ringdown is lasts longer for the cubic EFT of GR waveform and
with a smaller instantaneous frequency (see the bottom panel) due to the negative value of the
$\delta\omega^{(0)}$ coefficient in this theory.

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/example_waveform_cubicEFT.pdf}
\caption{(Color online).~Example GW signal with GW1509140-like parameters
for both GR (solid line) and cubic EFT of GR (dashed line) with the leading-order
$n=0$ modifications to the fundamental QNM.
%
The former is computed with the \SEOB{} model, while the latter with the
\pSEOB{} model, with ringdown modifications according to the results in
Table~\ref{tab:ref_theories_qnms}.
%
Top panel: the $+$ polarization $h_{+}(t)$. Bottom panel: the GW amplitude
$|\tilde{h}(t)|$ (left axes) and the instantaneous frequency $f(t)$ (right axes).
}
\label{fig:example_waveform}
\end{figure}

@@

Assuming Gaussian and stationary noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln P(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}\,,
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk denotes
complex conjugation and $S_{n}(f)$ is the power spectrum density of the
detector.
%
Assuming a specific prior distribution for our parameters, we stochastically
sample over the parameter space using a Markov-Chain Monte Carlo algorithm as implemented in
\texttt{LALInferenceMCMC}~\cite{Rover:2006ni,vanderSluys:2008qx},
which is part of the \texttt{LALInference} software suite~\cite{Veitch:2014wba,lalsuite}.
%
We subsequently marginalize over the remaining parameters to obtain the
posterior probability distribution function~(PDF) on $\ell$, our main quantity of interest.
%
% An example of this is shown in Fig. XX.

@@

% Let us show one concrete example of how this mapping is done.
%
In Ref.~\cite{Wagle:2021tam} the QNMs of slow-rotating BHs in dCS gravity were calculated
and it was found that the axial gravitational modes have their damping time increased as we
increase the Chern-Simons coupling $\ell_{\rm CS}$ at fixed spin $\chi$ of the BH.
%
Hence, according to the prescription of Sec.~\ref{sec:theory_specific_qnm}, this is the branch of QNMs we will select.

We can now proceed to determine $\delta\omega^{(n)}$ and $\delta\tau^{(n)}$ as follows.
%
Using the fitting formula Eq.~(54a) of~\cite{Wagle:2021tam}, namely,
%
\begin{equation}
    M \omega_{\rm CS} = c_{1} + c_{2} \kappa \zeta + (c_{3} + c_{4} \kappa \zeta) \, (1 - \chi)^{c_{5} + c_{6} \kappa \zeta},
    \label{eq:omega_dsc_eg}
\end{equation}
%
and similarly for the imaginary part ${\rm Im}(\sigma_{\rm CS}) =  - 1 / \tau_{\rm CS}$.
%
Here $\kappa = 1/(16 \pi)$, $\zeta = \ell_{\rm CS}^{4} / (M^4 \kappa)$, and thus
%
\begin{equation}
\kappa \zeta = (\ell_{\rm CS} / M)^4 \,,
\end{equation}
%
and $c_{i}$ are fitting coefficients which can be found in Ref.~\cite{Wagle:2021tam}, Table II.
%
In the context of a binary BH merger, $M$ is interpreted as the remnant mass $\Mf$.

We can now expand Eq.~\eqref{eq:omega_dsc_eg} to leading-orders in $\chi$ and $\ell_{\rm CS}$, and
gather the terms proportional to $\ell_{\rm CS}$.
%
We obtain
%
\begin{align}
    \Mf \, \omega^{\rm CS}_{220} &=
    \left[ 0.3722 + 1.1945 (\ell_{\rm CS} / \Mf)^4 \right]
    \nonumber \\
    &\quad + \left[0.1861 + 5.1828 (\ell_{\rm CS} / \Mf)^4 \right] \, \chi\,,
    \label{eq:omega_dcs_interm}
\end{align}
%
where we made use of the numerical value of the coefficients $c_{i}$.
%
We find that (reassuringly) that the nonrotating GR part of the expression above agree with $\omega^{(0)}$ of Ref.~\cite{Maselli:2019mjd}
to 0.5\% relative percent error. The same error estimate is larger ($\approx 20$\%) for the linear-in-spin coefficient;
0.1861 in comparison to 0.1258 of Ref.~\cite{Maselli:2019mjd}.
%
We attribute this difference to Ref.~\cite{Wagle:2021tam} having fitted Eq.~\eqref{eq:omega_dsc_eg} to QNM data compute to linear-order in spin, whereas~\cite{Maselli:2019mjd}
fitted Eq.~\eqref{eq:kerr_expansion} to Kerr QNM valid to all orders in spin.

We can now isolate the dCS corrections from Eq.~\eqref{eq:omega_dcs_interm} and
compare against Eq.~\eqref{eq:qnm_nongr_iso}, to find
%
\begin{equation}
p_{\rm CS} = 4\,, \quad \delta \omega^{(0)}_{\rm CS} = 3.1964\,, \quad \delta \omega^{(1)}_{\rm CS} = 41.199\,.
\label{eq:cs_omega_coefs}
\end{equation}

We can carry the same steps for $\tau_{\rm CS}$ and find
%
\begin{equation}
\delta \tau^{(0)}_{\rm CS} = 6.3619\,, \quad \delta \tau^{(1)}_{\rm CS} = 794.66\,.
\label{eq:cs_tau_coefs}
\end{equation}
%
which completes the set of fixed parameters beyond-GR parameters in the ringdown of the \pSEOB{} waveform model.
%
We remark that the alarmingly large values of $\delta \omega_{\rm CS}^{(1)}$ and $\delta \tau_{\rm CS}^{(1)}$
are compensated by the assumptions that $(\ell_{\rm CS} / M_f)^2$ and $\chi$ are much less than unity. These
are the assumptions use to calculate the QNM frequencies both in Ref.~\cite{Wagle:2021tam}.
